



<h1 style="color: #1E3A5F;"> Work Based Learning Assignments</h1>
<b>Author:</b> Jane Daisy Adhiambo<br>
<b>Objective:</b> learning computer science practical from Cordio East Africa<br>
<b>Mentors</b>: James Mbugua and Nimrod Ishmael

<body style="background-color:#E3E6E8;">
<h2 style="color: #E0B54C;">Mangrove Forest Degradation Analysis Based on Stump Count</h2>
<h4 style="color: #C9A9A6;">Breakdown of the Script:</h4>
<p><i>
The R script processes data from an Excel file (Mangrove_sample_data.xlsx) to analyze and visualize mangrove forest degradation based on the number of stumps recorded. It first loads necessary libraries (readxl, openxlsx, ggplot2, dplyr) for data manipulation and visualization. The script then groups the data by block names and forest types, summarizing the total number of stumps in each category. Two bar charts are generated: one displaying stump counts per block and another showing stump counts per forest type, helping to identify areas with higher degradation levels. The visualization uses a minimal theme with clear labels to enhance readability.<i>

</p>

</body>
<h5 style= "color: #4C7D87;">Libraries</h5>
```{r echo=TRUE, fig.align='center', fig.height=12, fig.width=12, message=FALSE, warning=FALSE}
library(readxl)
library(ggplot2)
library(openxlsx) 
library(dplyr)
```

<h5 style= "color: #4C7D87;">Summarize Stumps Per Block</h5>
```{r echo=TRUE, fig.align='center', fig.height=12, fig.width=12, message=FALSE, warning=FALSE}
data_excel <- "./Daisy_Challenge2/Mangrove_sample_data.xlsx" 
df <- read.xlsx(data_excel, sheet = 1)


df_summary <- df %>%
  group_by(Block_Names) %>%
  summarise(Total_Stumps = sum(Stumps, na.rm = TRUE))

print(df_summary)  
```
<h5 style= "color: #4C7D87;">bar graph presentation</h5>
```{r echo=TRUE, fig.align='center', fig.height=12, fig.width=12, message=FALSE, warning=FALSE}
ggplot(df_summary, aes(x = Block_Names, y = Total_Stumps, fill = Block_Names)) +
  geom_bar(stat = "identity") + #use actual values from df_summary
  labs(title = "Degradation by Stumps in Different Blocks",
       x = "Block Name",
       y = "Total Stumps (Degradation Level)") +
  theme_minimal() +
  theme(legend.position = "none")  # Remove legend since colors just help differentiate

```
<h2 style="color: #E0B54C;">Degradation by Stumps in Different Forest Types</h2>

<h5 style= "color: #4C7D87;">Summarize Stumps Per Block</h5>

```{r echo=TRUE, fig.align='center', fig.height=12, fig.width=12, message=FALSE, warning=FALSE}
data_excelsheet <- "./Daisy_Challenge2/Mangrove_sample_data.xlsx" 
df1 <- read.xlsx(data_excelsheet, sheet = 1)


df_summary1 <- df1 %>%
  group_by(Forest_Typ) %>%
  summarise(Total_Stumps = sum(Stumps, na.rm = TRUE))

print(df_summary1) 
```

<h5 style= "color: #4C7D87;">bar graph presentation</h5>
```{r echo=TRUE, fig.align='center', fig.height=12, fig.width=12, message=FALSE, warning=FALSE}
ggplot(df_summary1, aes(x = Forest_Typ, y = Total_Stumps, fill = Forest_Typ)) +
  geom_bar(stat = "identity") + #use actual values from df_summary1
  labs(title = "Degradation by Stumps in Different Blocks",
       x = "Forest_Typ",
       y = "Total Stumps (Degradation Level)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + # Rotate labels 45 degrees
  theme(legend.position = "none")  # Remove legend since colors just help differentiate
```

<h2 style="color: #E0B54C;">Gender-Based Attendance Analysis in Kilifi Central and Bofa</h2>

<p> <i>This R script visualizes gender-based attendance at two locations: Kilifi Central and Bofa. Using ggplot2, the script generates two bar charts representing the number of male and female attendees at each site. The data is structured in a simple dataframe, grouping attendees by gender. The Kilifi Central chart shows 18 male and 12 female attendees, while the Bofa chart displays 29 male and 12 female attendees. The bars are color-coded with blue for males and pink for females for easy differentiation. The visualization helps compare attendance patterns between the two locations, highlighting gender distribution. <i><p>



```{r}

library(ggplot2)

```


```{r pressure, echo=TRUE}
# Sample data: Number of attendees by gender
attendance_data <- data.frame(
  Gender = c("Male", "Female"),
  Count = c(18, 12)  # Replace with your actual numbers
)
ggplot(attendance_data, aes(x = Gender, y = Count, fill = Gender)) +
  geom_bar(stat = "identity",width = 0.3) +
  labs(title = "Kilifi central Attendance by Gender", x = "Gender", y = "Number of Attendees") +
  theme_minimal() +
  scale_fill_manual(values = c("Male" = "blue", "Female" = "pink"))


```
```{r, echo=TRUE}
# Sample data: Number of attendees by gender
attendance_data <- data.frame(
  Gender = c("Male", "Female"),
  Count = c(29, 12)  # Replace with your actual numbers
)
ggplot(attendance_data, aes(x = Gender, y = Count, fill = Gender)) +
  geom_bar(stat = "identity",width = 0.3) +
  labs(title = "Bofa  Attendance by Gender", x = "Gender", y = "Number of Attendees") +
  theme_minimal() +
  scale_fill_manual(values = c("Male" = "blue", "Female" = "pink"))


```

<h2 style="color: #E0B54C;">Mapping Mangrove Forest Distribution and Degradation in Lamu County</h2>
<p><i>This R script processes and visualizes mangrove forest data in Lamu County by converting Excel (.xlsx) data into CSV format, extracting latitude and longitude coordinates, converting them from Degrees, Minutes, Seconds (DMS) to Decimal Degrees (DD), and exporting the data as a shapefile (.shp). The script utilizes spatial libraries (sf, tmap, tmaptools) to create maps overlaying mangrove forest types onto the Lamu County boundary shapefile. The final visualization categorizes different forest types, displays their spatial distribution, and includes a compass, legend, and county label for better interpretation. The processed map is saved as an image for further analysis.<i></p>

<h5 style= "color: #4C7D87;">libraries</h5>
```{r echo= TRUE, fig.align='center', fig.height=12, fig.width=12, message=FALSE, warning=FALSE}
library(dplyr)    # Data manipulation
library(tidyr)    # Data tidying
library(stringr)  # String manipulation
library(sf) # Spatial data (Shapefile export)
library(sp) 
library(readr)    # Reading data (CSV, etc.)
library(bslib)
library(openxlsx) # for the excel
library(tmap)
library (tmaptools)#additional tools for maps
```

<h5 style= "color: #4C7D87;"> changing .xlsx to csv </h5>
```{r echo= TRUE, fig.align='center', fig.height=12, fig.width=12, message=FALSE, warning=FALSE}
# Correct file path
#excel_file <- "C:/Users/Cordio ea/Documents/Jane Daisy_Challenge/Mangrove_sample_data.xlsx"


excel_file<-"./Jane Daisy_Challenge/Mangrove_sample_data.xlsx"

# Read the first sheet of the Excel file
data_excel <- read.xlsx(excel_file, sheet = 1)
#data_locale <- "C:/Users/Cordio ea/Documents/Jane Daisy_Challenge/output_file.csv"
data_locale <- "./Jane Daisy_Challenge/output_file.csv"

# Save as CSV
write.csv(data_excel, data_locale, row.names = FALSE)


site_coordinates <- read.csv(data_locale)
# Display the first 5 rows of the data
head(site_coordinates, 5)
```

```{r echo=TRUE, fig.align='center', fig.height=12, fig.width=12, message=FALSE, warning=FALSE}

coordinate_extracted <- site_coordinates %>%
  mutate(
    # Extract Degrees, Minutes, and Seconds for Latitude
    Lat_degrees = as.numeric(str_extract(Lat, "^\\-?\\d+(?=째)")),  # Extract degrees (allow for negative)
    Lat_minutes = as.numeric(str_extract(Lat, "(?<=째)\\d+(?=')")),  # Extract minutes (after degree symbol)
   Lat_seconds = as.numeric(str_extract(Lat, "(?<=')\\d+(\\.\\d+)?(?=\")")),  # Extract seconds with decimal (after minutes)


    # Extract Degrees, Minutes, and Seconds for Longitude
    Long_degrees = as.numeric(str_extract(Lon, "^\\-?\\d+(?=째)")),  # Extract degrees (allow for negative)
    Long_minutes = as.numeric(str_extract(Lon, "(?<=째)\\d+(?=')")),  # Extract minutes (after degree symbol)
       Long_seconds = as.numeric(str_extract(Lon, "(?<=')\\d+(\\.\\d+)?(?=\")"))  # Extract seconds with decimal (after minutes)
  )

# Check the first few rows of the data after extraction
head(coordinate_extracted, 5)

```

```{r echo=TRUE, fig.align='center', fig.height=12, fig.width=12, message=FALSE, warning=FALSE}
coordinate_cleaned <- coordinate_extracted %>%
  mutate(
    # Clean Latitude and Longitude Degrees, Minutes, and Seconds
    Lat_degrees = as.numeric(Lat_degrees),  # Ensure degrees are numeric
    Lat_minutes = as.numeric(Lat_minutes),  # Ensure minutes are numeric
    Lat_seconds = as.numeric(Lat_seconds),  # Ensure seconds are numeric
    
    Long_degrees = as.numeric(Long_degrees),  # Ensure degrees are numeric
    Long_minutes = as.numeric(Long_minutes),  # Ensure minutes are numeric
    Long_seconds = as.numeric(Long_seconds),  # Ensure seconds are numeric
    
    # Adjust for negative values (if applicable)
    Lat_degrees = if_else(str_detect(Lat, "S"), -Lat_degrees, Lat_degrees),  # Handle Southern Hemisphere
    Long_degrees = if_else(str_detect(Lon, "W"), -Long_degrees, Long_degrees),  # Handle Western Hemisphere
    
    # Convert Latitude and Longitude to Decimal Degrees (DD)
    Lat_decimal = Lat_degrees + (Lat_minutes / 60) + (Lat_seconds / 3600),
    Long_decimal = Long_degrees + (Long_minutes / 60) + (Long_seconds / 3600)
  )

# Check the first few rows after cleaning and conversion
#head(coordinate_cleaned, 5)
# Save as CSV
write.csv(coordinate_cleaned, data_locale, row.names = FALSE)
DD_coordinates <- read.csv(data_locale)

# Display the first 5 rows of the data
head(DD_coordinates, 5)
```



```{r echo=TRUE, fig.align='center', fig.height=12, fig.width=12, message=FALSE, warning=FALSE}
# Convert to sf object (Spatial Data Frame)
sf_data <- st_as_sf(DD_coordinates, coords = c("Long_decimal", "Lat_decimal"), crs = 4326)


# Check the structure of the spatial data
print(sf_data)
#shapefile_path <- "C:/Users/Cordio ea/Documents/Jane Daisy_Challenge/output_shapefile.shp"
shapefile_path <-"./Jane Daisy_Challenge/output_shapefile.shp"
st_write(sf_data, shapefile_path, delete_layer = TRUE)



```
<h5 style= "color: #4C7D87;">Steps to Create a Map with Categories</h5>
```{r echo=TRUE, fig.align='center', fig.height=12, fig.width=12, message=FALSE, warning=FALSE}

# Load the shapefile
#sf_map <- st_read("shapefile_path")

# Check column names
#colnames(sf_map)

# Create the map and assign it to an object
#mangrove_map <- tm_shape(sf_map) + 
 # tm_dots(col = "Vegettn", palette = "brewer.set1", title = "Vegetation Type") + 
 # tm_basemap("OpenStreetMap") + 
  #tm_layout(legend.outside = TRUE)

# Save the map as an image
#tmap_save(mangrove_map, filename = "C:/Users/Cordio ea/Documents/Jane Daisy_Challenge/mangrove_map.png")


```

```{r echo=TRUE, fig.align='center', fig.height=12, fig.width=12, message=FALSE, warning=FALSE}

# Load the mangrove points (output_shapefile)
#sf_mangrove <- st_read("C:/Users/Cordio ea/Documents/Jane Daisy_Challenge/output_shapefile.shp")
sf_mangrove <-st_read("./Jane Daisy_Challenge/output_shapefile.shp")


# Load Lamu County shapefile
#lamu_boundary <- st_read("C:/Users/Cordio ea/Documents/Jane Daisy_Challenge/Lamu_County.shp")
lamu_boundary <- st_read("./Jane Daisy_Challenge/Lamu_County.shp")




# Ensure both layers have the same CRS
sf_mangrove <- st_transform(sf_mangrove, crs = 4326)
lamu_boundary <- st_transform(lamu_boundary, crs = 4326)

# Check column names to confirm "Forest_Typ"
print(colnames(sf_mangrove))  # Ensure "Forest_Typ" exists

# Create the overlay map
lamu_map <- tm_shape(lamu_boundary) + 
  tm_borders(lwd = 2, col = "black", title = "Lamu County Boundary") +  # Lamu County Outline
  tm_text("Lamu ", size = 1.0, col = "black", fontface = "bold", xmod = -0.15, ymod = -0.05) +        # Label the County Name
  
  tm_shape(sf_mangrove) + 
  tm_dots(col = "Frst_Ty", palette = "brewer.set1", size = 0.5, title = "Forest Type") +  # Forest Points
# Add a compass in the top-left corner
  tm_compass(position = c("right", "bottom"), type = "8star", size = 2) +
  # Add a title to the map
  tm_layout(title = "Forest Type Distribution in Lamu County", 
            title.size = 1.1,
            legend.outside = TRUE) 
  
  tm_layout(legend.outside = TRUE)

# Show the map
lamu_map

# Save the map
#tmap_save(lamu_map, filename = "C:/Users/Cordio ea/Documents/Jane Daisy_Challenge/lamu_mangrove_map.png")
tmap_save(lamu_map, filename = "./Jane Daisy_Challenge/lamu_mangrove_map.png")
#
#
```
<h2 style="color: #E0B54C;">Processing and Analyzing Sea Surface Temperature (SST) Data for Spatial Mapping</h2>


<p><i>This R script processes Sea Surface Temperature (SST) data from a CSV file by converting it into a spatial format and generating a shapefile (.shp) for mapping. The script first reads the raw site data, converts latitude and longitude into a Simple Features (sf) object, and exports it as a shapefile for spatial analysis. Then, it processes SST data by transforming it into long format, renaming date columns, and converting SST values from Kelvin to Celsius. It further calculates summary statistics such as maximum, minimum, and standard deviation of SST values. Finally, the processed data is saved as a CSV file for further analysis and visualization.<i></p>

```{r echo= TRUE, fig.align='center', fig.height=12, fig.width=12, message=FALSE, warning=FALSE}
library(sf)
library(sp)
library(tidyverse)

```


```{r echo= TRUE, fig.align='center', fig.height=12, fig.width=12, message=FALSE, warning=FALSE }
# Read the CSV file
csv_file <- "C:/Users/Cordio ea/Documents/JaneDaisy_WorkSpace/EOData_linking/Raw_sites.csv"
data <- read.csv(csv_file)

# Convert to sf object (Simple Features)
data_sf <- st_as_sf(data, coords = c("lon", "lat"), crs = 4326)  # WGS84 CRS

```


```{r echo= TRUE, fig.align='center', fig.height=12, fig.width=12, message=FALSE, warning=FALSE}
# Define output path
output_folder <- "C:/Users/Cordio ea/Documents/JaneDaisy_WorkSpace/EOData_linking/"
output_file <- paste0(output_folder, "Raw_sites.shp")

# Write the Shapefile
st_write(data_sf, output_file, delete_layer = TRUE)

cat("Shapefile saved at:", output_file)
```
```{r}
extracted_data <- read.csv("C:/Users/Cordio ea/Documents/JaneDaisy_WorkSpace/EOData_linking/SST_Extracted.csv")
colnames(extracted_data)
```

<h5 style= "color: #4C7D87;">Converting SST to long format</h5>

```{r echo= TRUE, fig.align='center', fig.height=12, fig.width=12, message=FALSE, warning=FALSE}


extracted_data_long <- extracted_data %>%
  pivot_longer(
    cols = starts_with("X"),  # Select all columns starting with "X" (date columns)
    names_to = "Date", 
    values_to = "SST"
  )
extracted_data_long <- extracted_data_long %>%
  mutate(Date = gsub("^X", "", Date)) %>%  # Remove "X"
  mutate(Date = as.Date(Date, format="%d_%b_%Y"))  # Convert to Date format

head(extracted_data_long)


```

```{r echo= TRUE, fig.align='center', fig.height=12, fig.width=12, message=FALSE, warning=FALSE}
# Convert SST column to numeric
extracted_data_long$SST <- as.numeric(extracted_data_long$SST)

# Convert from Kelvin to Celsius using SST column
extracted_data_long <- extracted_data_long %>%
  mutate(SST_Celsius = SST - 273.15)

# Compute max, min, and standard deviation for SST in Celsius
max_temp_Celsius <- max(extracted_data_long$SST_Celsius, na.rm = TRUE)
min_temp_Celsius <- min(extracted_data_long$SST_Celsius, na.rm = TRUE)
std_dev_temp_Celsius <- sd(extracted_data_long$SST_Celsius, na.rm = TRUE)

# Add computed values as new columns in Celsius
extracted_data_long <- extracted_data_long %>%
  mutate(
    max_temp = max_temp_Celsius,   # Now in Celsius
    min_temp = min_temp_Celsius,  
    std_dev_temp = std_dev_temp_Celsius  # Now in Celsius
  )

# Save the processed dataset to CSV
write.csv(extracted_data_long, "C:/Users/Cordio ea/Documents/JaneDaisy_WorkSpace/EOData_linking/SST_Processed.csv", row.names = FALSE)

# Print a short summary
head(extracted_data_long)

```
```{r}
View(extracted_data_long)


```

<h2 style="color: #E0B54C;">Summary statistics and histograms for WLR and SLA_Interp</h2>

```{r echo= TRUE, fig.align='center', fig.height=12, fig.width=12, message=FALSE, warning=FALSE}
library(ggplot2)
library(dplyr)
library(moments)  # For kurtosis calculation
library(readr)    # For reading CSV files

```

```{r echo= TRUE, fig.align='center', fig.height=12, fig.width=12, message=FALSE, warning=FALSE}
df <- read_csv("C:/Users/Cordio ea/Documents/Jane Daisy_Challenge/Raw_Ke_Transects_rates_sla.csv")
str(df)  # Check column names and types
summary(df)  # Quick summary of numerical values


```

```{r echo= TRUE, fig.align='center', fig.height=12, fig.width=12, message=FALSE, warning=FALSE}
df <- df[!is.na(df$WLR) & !is.na(df$SLA_Interp), ]



```

```{r echo= TRUE, fig.align='center', fig.height=12, fig.width=12, message=FALSE, warning=FALSE}
# Compute summary statistics
summary_stats <- data.frame(
  Statistic = c("Mean", "Standard Deviation", "Kurtosis", "Maximum"),
  WLR = c(mean(df$WLR), sd(df$WLR), kurtosis(df$WLR), max(df$WLR)),
  SLA_Interp = c(mean(df$SLA_Interp), sd(df$SLA_Interp), kurtosis(df$SLA_Interp), max(df$SLA_Interp))
)

# Print the summary statistics
print(summary_stats)
```
```{r }
# Create histogram for WLR with log-transformed frequency and x-axis starting at -30
wlr_log_plot <- ggplot(df, aes(x = WLR)) +
  geom_histogram(binwidth = 0.1, fill = "blue", color = "white", alpha = 0.7) +
  scale_y_log10() +  # Apply log transformation to y-axis
  xlim(-30, max(df$WLR, na.rm = TRUE)) +  # Ensure x-axis starts at -30
  labs(title = "Histogram of WLR (Log-Scaled Frequency)", x = "WLR Values", y = "Log(Frequency)") +
  theme_minimal() +  
  theme(
    plot.title = element_text(color = "white", size = 14, face = "bold"),
    axis.title.x = element_text(color = "white", size = 12),
    axis.title.y = element_text(color = "white", size = 12),
    axis.text = element_text(color = "white"),
    panel.background = element_rect(fill = "black", color = NA), # Black background
    plot.background = element_rect(fill = "black", color = NA)   # Black background
  )  

print(wlr_log_plot)


# Save the log-transformed histogram plot
#ggsave("C:/Users/Cordio ea/Documents/WLR_Histogram_Log.png", plot = wlr_log_plot, dpi = 300, width = 8, height = 6)



```


```{r }

# Create histogram for SLA_Interp with log y-axis
sla_plot <- ggplot(df, aes(x = SLA_Interp)) +
  geom_histogram(binwidth = 0.05, fill = "red", color = "black", alpha = 0.7) +
  scale_y_log10() +  # Logarithmic Y-axis
  labs(title = "Histogram of SLA_Interp (Log-Scaled Frequency)", x = "SLA_Interp Values", y = "Log(Frequency)") +
  theme_minimal() +
  theme(
    plot.title = element_text(color = "white", size = 14, face = "bold"),
    axis.title.x = element_text(color = "white", size = 12),
    axis.title.y = element_text(color = "white", size = 12),
    axis.text = element_text(color = "white"),
    panel.background = element_rect(fill = "black", color = NA), # Black background
    plot.background = element_rect(fill = "black", color = NA)   # Black background
  )

print(sla_plot)


```




